Location and Scale-Invariant Power Transformations for Transforming Data to

### 2025-04-08 manuscript_rmarkdown/manuscript.Rmd
Update experimental results.

A common feature in health-related datasets is age.
Here we use data on 228 patients with lung cancer that was collected and published by Loprinzi et al. [@Loprinzi1994-cd].
The age in the cohort was $62.4 \pm 9.1$ (mean ± standard deviation) years.
Applying conventional and invariant Yeo-Johnson transformations to patient age yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (sum of residuals with normal distribution $\sum r_i = 16.5$, $p = 0.71$);
conventional transformation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$, $p = 0.94$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$, $p = 0.94$);
location- and scale-invariant transformation ($\lambda = 0.9$, $\sum r_i = 8.8$, $\mu_{YJ} = 1.2$, $\sigma_{YJ} = 1.1$, $p = 0.95$);
and robust location- and scale-invariant transformation ($\lambda = 0.8$, $\sum r_i = 10.6$, $\mu_{YJ} = 1.2$, $\sigma_{YJ} = 1.0$, $p = 0.80$).

Location- and scale-invariant transformation led to a lower overall residual sum, indicating a better transformation.
Robust location- and scale-invariant transformation had a higher residual sum compared to the non-robust variant, which may be due to the lack of outliers in the data.
Conventional transformations inflated the mean $\mu_{YJ}$ and standard deviation $\sigma_{YJ}$ of the age feature after transformation.
The empirical central normality test did not detect any statistically significant deviations from central normality for any transformation (all $p \geq 0.78$).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342 penguins of three different species [@Gorman2014-eo].
The body mass was $(4.2 \pm 0.8) \cdot 10^3$ (mean ± standard deviation) grams, and not centrally normal ($p = 0.03$).
Applying conventional and invariant Yeo-Johnson transformations to body mass yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (residual sum $\sum r_i = 48.0$, $p < 0.001$);
conventional transformation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$, $p = 0.06$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$, $p = 0.06$);
location- and scale-invariant transformation ($\lambda = 0.5$, $\sum r_i = 26.8$, $\mu_{YJ} = 0.9$, $\sigma_{YJ} = 0.9$, $p = 0.18$);
and robust location- and scale-invariant transformation ($\lambda = 0.4$, $\sum r_i = 23.1$, $\mu_{YJ} = 0.8$, $\sigma_{YJ} = 0.8$, $p = 0.40$).

Location- and scale-invariant transformation produced a lower overall residual sum, indicating a better transformation.
Moreover, conventional transformations led to low standard deviation $\sigma_{YJ}$ of the body mass feature after transformation.
The empirical central normality test did not detect any statistically significant deviations from central normality for any transformation (all $p \geq 0.38$).

@@

Geospatial datasets usually contain coordinates.
The Ames housing dataset contains data on 2930 properties that were sold between 2006 and 2010 [@De-Cock2011-jf], including their geospatial coordinates.
The latitude was $42.03 \pm 0.02)$ (mean ± standard deviation).
Applying conventional and invariant Yeo-Johnson transformations to latitude yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (residual sum $\sum r_i = 328$, $p < 0.001$);
conventional transformation ($\lambda = 62.1$, $\sum r_i = 319$, $\mu_{YJ} = 4.8 \cdot 10^{99}$, $\sigma_{YJ} = 0.1 \cdot 10^{99}$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 95.4$, $\sum r_i = 319$, $\mu_{YJ} = 6.4 \cdot 10^{153}$, $\sigma_{YJ} = 0.3 \cdot 10^{153}$, $p < 0.001$);
location- and scale-invariant transformation ($\lambda = 1.5$, $\sum r_i = 326$, $\mu_{YJ} = -1.2$, $\sigma_{YJ} = 0.8$, $p < 0.001$);
and robust location- and scale-invariant transformation ($\lambda = 1.4$, $\sum r_i = 311$, $\mu_{YJ} = -1.3$, $\sigma_{YJ} = 0.9$, $p < 0.001$).

Every transformation reduced the residual sum.
The non-robust location- and scale-invariant transformation did not improve over conventional alternatives and yielded a data distribution lacking central normality (empirical central normality test: $p=0.05$).
However, conventional transformations had high values for the $\lambda$ parameter, which could lead to numerical issues.

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the BBC television show *Top Gear* [@Alfons2021-kc].
Within this dataset, the fuel consumption feature contains outliers due to highly fuel-efficient vehicles.
Applying conventional and invariant Yeo-Johnson transformations to the fuel consumption feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 54$, $p=0.58$);
conventional transformation ($\lambda = -0.1$, $\sum r_i = 55$, $\mu_{YJ} = 3.0$, $\sigma_{YJ} = 0.3$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 0.8$, $\sum r_i = 48$, $\mu_{YJ} = 29$, $\sigma_{YJ} = 15$, $p=0.32$);
location- and scale-invariant transformation ($\lambda = -1.3$, $\sum r_i = 44$, $\mu_{YJ} = 0.5$, $\sigma_{YJ} = 0.1$, $p < 0.001$);
and robust location- and scale-invariant transformation ($\lambda = -0.9$, $\sum r_i = 50$, $\mu_{YJ} = 2.0$, $\sigma_{YJ} = 2.3$, $p=0.35$).

@@

The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
These patients underwent Computed Tomography Angiography to characterize the carotid artery blockages.
Angiography imaging was then assessed, and various characteristics related to the blood vessels and the disease are measured.
The maximum arterial wall thickness feature contains several instances with outlier values.
Applying conventional and invariant Yeo-Johnson transformations to this feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 110$, $p=0.86$);
conventional transformation ($\lambda = -0.7$, $\sum r_i = 30$, $\mu_{YJ} = 1.0$, $\sigma_{YJ} = 0.1$, $p=0.02$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 1.1$, $\sum r_i = 136$, $\mu_{YJ} = 7.2$, $\sigma_{YJ} = 14$, $p=0.91$);
location- and scale-invariant transformation ($\lambda = 0.2$, $\sum r_i = 12$, $\mu_{YJ} = -11.8$, $\sigma_{YJ} = 6.9$, $p=0.23$);
and robust location- and scale-invariant transformation ($\lambda = -0.6$, $\sum r_i = 27$, $\mu_{YJ} = 0.7$, $\sigma_{YJ} = 0.1$, $p=0.18$).

The conventional non-robust transformation failed to produce a centrally normal distribution (empirical central normality test $p=0.02$).
Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).

>>>>

A common feature in health-related datasets is age.
Here we use data on 228 patients with lung cancer that was collected and published by Loprinzi et al. [@Loprinzi1994-cd].
The age in the cohort was $62.4 \pm 9.1$ (mean ± standard deviation) years.
Applying conventional and invariant Yeo-Johnson transformations to patient age yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation ($\sum r_i = 16.5$, $p = 0.69$);
conventional transformation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$, $p = 0.96$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$, $p = 0.96$);
location- and scale-invariant transformation ($\lambda = 1.3$, $\sum r_i = 8.8$, $\mu_{YJ} = -1.2$, $\sigma_{YJ} = 1.1$, $p = 0.98$);
and robust location- and scale-invariant transformation ($\lambda = 1.3$, $\sum r_i = 9.3$, $\mu_{YJ} = -1.0$, $\sigma_{YJ} = 1.1$, $p = 0.93$).

Location- and scale-invariant transformation led to a lower overall residual errors, indicating a better transformation.
Conventional transformations inflated the mean $\mu_{YJ}$ and standard deviation $\sigma_{YJ}$ of the age feature after transformation.
The empirical central normality test did not detect any statistically significant deviations from central normality for any transformation (all $p \geq 0.93$).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342 penguins of three different species [@Gorman2014-eo].
The body mass was $(4.2 \pm 0.8) \cdot 10^3$ (mean ± standard deviation) grams, and not centrally normal ($p < 0.001$).
Applying conventional and invariant Yeo-Johnson transformations to body mass yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (residual sum $\sum r_i = 48.0$, $p < 0.001$);
conventional transformation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$, $p = 0.10$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$, $p = 0.10$);
location- and scale-invariant transformation ($\lambda = 0.5$, $\sum r_i = 26.8$, $\mu_{YJ} = 0.9$, $\sigma_{YJ} = 0.9$, $p = 0.28$);
and robust location- and scale-invariant transformation ($\lambda = 0.3$, $\sum r_i = 22.0$, $\mu_{YJ} = 0.7$, $\sigma_{YJ} = 0.9$, $p = 0.69$).

Location- and scale-invariant transformation produced a lower overall residual errors, indicating a better transformation.
Moreover, conventional transformations led to low standard deviation $\sigma_{YJ}$ of the body mass feature after transformation.
The empirical central normality test did not detect any statistically significant deviations from central normality for any transformation (all $p \geq 0.10$).

@@

Geospatial datasets usually contain coordinates.
The Ames housing dataset contains data on 2930 properties that were sold between 2006 and 2010 [@De-Cock2011-jf], including their geospatial coordinates.
The latitude was $42.03 \pm 0.02)$ (mean ± standard deviation).
Applying conventional and invariant Yeo-Johnson transformations to latitude yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (residual sum $\sum r_i = 328$, $p < 0.001$);
conventional transformation ($\lambda = 62.1$, $\sum r_i = 319$, $\mu_{YJ} = 4.8 \cdot 10^{99}$, $\sigma_{YJ} = 0.1 \cdot 10^{99}$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 95.4$, $\sum r_i = 315$, $\mu_{YJ} = 6.4 \cdot 10^{153}$, $\sigma_{YJ} = 0.3 \cdot 10^{153}$, $p < 0.001$);
location- and scale-invariant transformation ($\lambda = 1.5$, $\sum r_i = 326$, $\mu_{YJ} = -1.2$, $\sigma_{YJ} = 0.8$, $p < 0.001$);
and robust location- and scale-invariant transformation ($\lambda = 1.1$, $\sum r_i = 308$, $\mu_{YJ} = -1.3$, $\sigma_{YJ} = 1.2$, $p < 0.001$).

Every transformation reduced the residual sum.
None of the transformations yielded a centrally normal distribution.
Conventional transformations had high values for the $\lambda$ parameter, which could lead to numerical issues.

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the BBC television show *Top Gear* [@Alfons2021-kc].
Within this dataset, the fuel consumption feature contains outliers due to highly fuel-efficient vehicles.
Applying conventional and invariant Yeo-Johnson transformations to the fuel consumption feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 54$, $p=0.72$);
conventional transformation ($\lambda = -0.1$, $\sum r_i = 55$, $\mu_{YJ} = 3.0$, $\sigma_{YJ} = 0.3$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 0.8$, $\sum r_i = 48$, $\mu_{YJ} = 29$, $\sigma_{YJ} = 15$, $p=0.37$);
location- and scale-invariant transformation ($\lambda = -1.3$, $\sum r_i = 44$, $\mu_{YJ} = 0.5$, $\sigma_{YJ} = 0.1$, $p < 0.001$);
and robust location- and scale-invariant transformation ($\lambda = 1.0$, $\sum r_i = 56$, $\mu_{YJ} = 1.7$, $\sigma_{YJ} = 1.0$, $p=0.76$).

@@

The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
These patients underwent Computed Tomography Angiography to characterize the carotid artery blockages.
Angiography imaging was then assessed, and various characteristics related to the blood vessels and the disease are measured.
The maximum arterial wall thickness feature contains several instances with outlier values.
Applying conventional and invariant Yeo-Johnson transformations to this feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 110$, $p=0.83$);
conventional transformation ($\lambda = -0.7$, $\sum r_i = 30$, $\mu_{YJ} = 1.0$, $\sigma_{YJ} = 0.1$, $p=0.003$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 1.1$, $\sum r_i = 136$, $\mu_{YJ} = 7.2$, $\sigma_{YJ} = 14.3$, $p=0.88$);
location- and scale-invariant transformation ($\lambda = 0.2$, $\sum r_i = 12$, $\mu_{YJ} = -11.8$, $\sigma_{YJ} = 6.9$, $p=0.15$);
and robust location- and scale-invariant transformation ($\lambda = -0.3$, $\sum r_i = 30$, $\mu_{YJ} = 0.8$, $\sigma_{YJ} = 0.2$, $p=0.18$).

The conventional non-robust transformation failed to produce a centrally normal distribution (empirical central normality test $p=0.003$).
Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).
